---
title: 'Part 2: chi-squared with Lizards'' tails'
author: "Alex Setmajer"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(broom)
```

## Read in data
```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```

Get counts of lizards by species:
```{r}
lizards %>% 
  count(spp) %>% 
  arrange(-n)
```

## Make a subset

- UTST Lizards
- Sites: IBPE (enclosure grassland site), RABB (rabbit mesquite site), EAST (east tarbush site)
- Tail condition is recorded (W= whole, B= broken)

```{r}
utst <- lizards %>% 
  filter (spp == "UTST") %>% 
  filter (site %in% c("IBPE", "RABB", "EAST")) %>% 
  filter (tail %in% c("B", "W"))
```

## Find counts and proportions of lizards with broken or whole tails at each site

```{r}
utst %>% 
  count(site, tail)
```

Create a nice contigency table:
```{r}
tail_counts <- utst %>% 
  janitor::tabyl(site, tail) #creates table better for chi squared and communication, vs longform list above which is better for most data analysis/coding.

tail_proportions <- tail_counts %>% 
  janitor::adorn_percentages() %>% 
  janitor::adorn_pct_formatting(digits= 2) %>% 
  janitor::adorn_ns()
```

## Convert to a contigency table

Use `column_to_rowname()`.

```{r}
lizard_ct <- tail_counts %>% 
  column_to_rownames(var = "site") # site column no longer stored as data, now it is the row name.

```

## Chi-square

Are tail breakage and site independent?

```{r}
lizard_x2 <- chisq.test(lizard_ct)
lizard_x2

lizard_tidy <- broom::tidy(lizard_x2)
```

df = degrees of freedom is rows -1 * columns -1. p-value is probability in this test. below 0.05 so yes, have enough evidence to reject null hypothesis. AKA there is a significant association between site and tail breakage (lizard and tail breakage are NOT independent)

There is a significant association (i.e. non-independence) between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`) = `r lizard_tidy$statistic`, *p* = `r  lizard_tidy$p.value`).
